---
title: "Mortality in South African Provinces"
runtime: shiny
output: 
  flexdashboard::flex_dashboard:
    orientation: columns
    vertical_layout: fill
---

```{r setup, include=FALSE}
library(flexdashboard)
library(shiny)
library(dplyr)
library(tidyr)
library(ggplot2)
library(leaflet)
library(southafricastats)

totals <- population_zaf %>% 
    filter(year == 2013) %>% 
    select(province, total)

compare_provinces <- mortality_zaf %>%
    left_join(population_zaf) %>%
    filter(!is.na(total)) %>%
    mutate(mortality = deaths / total * 1e3) %>%
    group_by(province, indicator) %>%
    summarise(mortality = mean(mortality, na.rm = TRUE)) %>%
    ungroup %>%
    left_join(totals) %>%
    spread(indicator, mortality)

mortality_zaf <- mortality_zaf %>%
    filter(!(indicator %in% c("All causes")))

```

Column {.sidebar}
=====================================================================

This app explores mortality data from the [South Africa Data Portal](http://southafrica.opendataforafrica.org/).

Use the input below to select a cause of death to explore.

```{r}
selectInput(inputId = "indicator",
            label = "Cause of death:",
            choices = unique(mortality_zaf$indicator),
            selected = "Tuberculosis (A15-A19)")
```



Map {data-icon="fa-map-marker"}
====================================================================

### Which provinces have a higher mortality rate from the selected cause of death?

```{r}
selected_df <- reactive({
    subset_df <- compare_provinces[, c(1, 
                                       which(colnames(compare_provinces) == input$indicator))]
    colnames(subset_df) <- c("province", "indicator")
    population_zaf %>%
        filter(year == 2013) %>%
        left_join(subset_df, by = "province") %>%
        mutate(indicator = indicator / sum(indicator, na.rm = TRUE))
})

renderLeaflet({
    leaflet(selected_df()) %>%
        addProviderTiles("CartoDB.Positron") %>%
        addCircles(lng = ~longitude, lat = ~latitude, weight = 2.5,
             radius = ~sqrt(indicator) * 3e5 , popup = ~province,
             color = "magenta")
})
```

Comparing provinces {data-icon="fa-list"}
====================================================================

### How does the selected mortality rate compare to the overall mortality rate?

```{r}
scatterplot_df <- reactive({
    subset_df <- compare_provinces[, c(1:2, 
                                       which(colnames(compare_provinces) == "All causes"),
                                       which(colnames(compare_provinces) == input$indicator))]
    colnames(subset_df) <- c("province", "total", "selected_x", "selected_y")
    subset_df
})

renderPlot({
    ggplot(scatterplot_df(), aes(x = selected_x,
                                 y = selected_y,
                                 size = total,
                                 label = province)) +
        geom_point(alpha = 0.7, color = "magenta4") +
        theme_minimal() +
        labs(x = "All causes", y = input$indicator) +
        geom_text(aes(size = 1e7), vjust = 2) +
        scale_x_continuous(limits = c(7.7, 14.2)) +
        theme_minimal(base_size = 14) +
        theme(legend.position="none")
})
```

Changes in time {data-icon="fa-area-chart"}
====================================================================

### How have the number of deaths changed in time?

```{r}
renderPlot({
    mortality_zaf %>%
        filter(indicator == input$indicator) %>%
        ggplot(aes(year, deaths, color = province)) +
        geom_line(alpha = 0.8, size = 1.5) +
        theme_minimal(base_size = 18) +
        labs(x = NULL, y = "Number of deaths per year")
})
```


Table {data-icon="fa-table"}
====================================================================

### Explore the data as reported by the South Africa Data Portal

```{r}
renderDataTable({
    mortality_zaf %>%
        filter(indicator == input$indicator)
    },
    options = list(pageLength = 10)
    )
```
